Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-30 AnalyticStructure_main.tex
implemented minus corrections and clean-up of the tex. Update of the appendix is still missing

We will see in what follows that the situation is not as simple in the context of LR. Indeed we will show that the response density -- even for molecular systems -- looses its localized behaviour for high frequencies, independently on the nature of the specific system and perturbation. Furthermore, we will demonstrate that to increase the basis set completeness, delocalized oscillatory degrees of freedom cannot be excluded \textit{a priori}, even in the localized regime.

We will also see that the reliabilty of the convergence criteria that can be established for LR -- where no variational theorem exists -- depends explicitly on the target frequency range and on the features of the numerical basis set.

% This analysis evidences that the description of molecular systems beyond the ground state unavoidably leads to the inclusion of delocalized
% degrees of freedom above a specific energetic threshold. Moreover on the basis of this analysis we can assess something about the reliability of the results provided by computational setups able to explicit include the presence of delocalized degrees of freedom compared
% to the ones in which these degrees of freedom are excluded by design.

@@

From equation \eqref{rhoPrimeFluctuationStateDef2} we see that
%is the convenient starting point to assess the localization properties of the FS as it is easy to see that,
for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is spatially localized, \textit{i.e.} its evaluation can be restricted on a bounded real-space domain.
Even when such locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strenghts $F_j$ small enough to
preserve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
% \begin{multline}\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \\
% =2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
% \end{multline}
here we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime},
realized for $\omega>|\eps_h|$, in which fluctuation states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2}
can be evaluated in a localized domain. In this case a computational description of fluctuation states is only meaningful in ``weak''
sense, where the bound state character of $\bra{\psi_p\op O}$ behaves as a regulator. This is naturally achieved in the above mentioned
Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design}
a localized behaviour of the FS might reveal very dangerous in view of convergence of the results since the computational basis set has
to be able to express a delocalized FS in the support domain of $\bra{\psi_p\op O}$ .
This fact implies that the degrees of completeness of the computational basis should be increased by adding
the oscillatory degrees of freedom needed to describe the FS in the relevant domain.
Stated otherwise, to achieve precise results in this regime the
underlying basis set has to be able to express delocalized states.
%The degrees of completeness of the computational basis should therefore be increased by imposing oscillatory degrees of freedom needed to
%represent the FS at the border of the simulation domain: it is important that the underlying basis set is able to express delocalized states.

@@

It is possible to reformulate this type of analysis by directly looking at LR quantities that are defined independently from the perturbing operators and target observable. This can be achieved by studying the \emph{susceptibility functional}, defined implicitly through the operator equation
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \left(\omega -\mathcal L \right)^{-1} \commutator{\op \Phi(\omega)}{\dmnot} = \sop{{\cal \chi}}(\omega) \op\Phi(\omega) \;.
\ee
The susceptibility (super)operator $\cal \chi(\omega)$ is therefore closely related to the resolvent of the Liouvillian of Eq.~\eqref{LiouvillianRhopomegaDef1}, which is completely defined
from the molecule Hamiltonian $\hnot[\dm]$.
The ``excitation modes''  of the molecule are defined through the \emph{excitation operators}, satisfying
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$.
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) = \sum_{\{a\}} \,
\frac{\sop B_a}{\omega-\Omega_a} \;,
\ee
where the action of the spectral (super)operator $\sop B_a$ reads
\be\lb{spectralop}
  \sop B_a \cdot = \op E_a \trace{\commutator{\dmnot}{\op{\tilde E}_a} \cdot} \;. \nn
\ee
In this way Eq.~\eqref{LinearResponseFunctDef2} may be also written as:
\be\lb{deltaoexc}
  \braket{\delta \op O} =  \sum_{\{a\}} \frac{\trace{\op O \sop B_a \op \Phi(\omega)}}{\omega -\Omega_a}\;,
\ee

% \be
%   \trace{\op O \sop B_a \op \Phi(\omega)} = \trace{\op O \op E_a}
%   \trace{\op{\tilde E}_a \commutator{\dmnot}{\op \Phi(\omega)}} \;. \nn
% \ee

@@

To further address this point we exploit the formal relation between response density and susceptibility functional %established in ~\eqref{SusceptibilityDef1}
$\dm'(\omega) = \sop \chi(\omega)\op\Phi(\omega)$ and make usage of the representation ~\eqref{rhoPrimeFluctuationStateDef1} of the response density. Pursuing this procedure provides the excitation-based
representation of FS as follows:
% Let us now consider the features of the response density operator of Eq.~\eqref{rhoPrimeFluctuationStateDef1}.
% The fluctuation state in this case can be written as
% \be\lb{fsinexc}
% \ket{f_p^{\mathbf r_j}(\omega)}=
% \sop \chi(\omega) \op{\mathbf r_j} \ket{\psi_p}
% = \sum_a \frac{\bra{w_p^a} \op{\mathbf r_j} \ket{\psi_p}}{\omega^2 - \Omega_a^2}
% \ket{w_p^a}\;,
% \ee
\be\lb{fsinexc}
\ket{f_p^{\mathbf r_j}(\omega)}=
\sum_a \frac{\bra{w_p^a} \op{\mathbf r}_j \ket{\psi_p}}{\omega^2 - \Omega_a^2}
\ket{w_p^a}\;,
\ee
where we have employed the notation $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
This is the well-known equivalence of the Sternheimer formalism with the spectral representation of the LR susceptibility,
that is always valid in any computational basis set provided the the full spectrum of excited states (including the pseudo-continuum ones) is considered.
%The above equation shows that for \emph{any} value of $\omega$ both sectors of localized and delocalized states $\ket{w_p^a}$ contribute.
The above equation shows that both the sectors of localized and delocalized states $\ket{w_p^a}$ contribute to the FS for \emph{any} value of $\omega$.
However the discussion of Sec.~\ref{FluctuationState} has shown that the FS is a genuinely localized state below the IP threshold.
The equivalence implies that in this regime the contribution of the delocalized excitations in Eq.~\eqref{fsinexc} has to resum into a localized, asymptotically vanishing FS.
% Such a locality is evident for the imaginary part of the FS, which can be straightforwardly restricted to the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). For the real part of FS the \emph{entire} spectrum
% of excitations contribute.
The localization below threshold is manifest for the imaginary part of the FS, that is explicitly written as the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). Instead, for the real part of FS, the \emph{entire} excitation spectrum contributes to the construction of the FS,
which has to be interpreted as a localized wavepacket in the basis of the pseudo-continuum excitations.
% and the localization
%\emph{has} to realize due to a non trivial cancellation of the unbound parts of the $\ket{w_p^a}$'s.

@@

\subsection{Effective representation of $\sop \chi(\omega)$ for localized response densities}
The above arguments have shown that the only quantity which can be expressed in terms of the localized sector of the excitation spectrum is the imaginary part of the susceptibility functional $\sop \chi(\omega)$ in the optical regime (i.e. $\omega <$  IP). The response density for a generic perturbation and $\omega$ value requires the inclusion in the Liouvillian DOS of the delocalized part of the excitation spectrum, even in view of expressing localized fluctuation states.

A given numerical treatment typically provides only an effective description of the sector of delocalized excitation which in turns gives rise to an effective
representation of the susceptibility functional. It appears therefore important to identify indicators that enable us to validate the capability
of $\sop \chi(\omega)$ to express the localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
%It appears therefore important to identify indicators that might assess the capability of
%$\sop \chi(\omega)$ to express localised FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
For instance, the static polarizability tensor $\alpha_{ij}$ can be calculated as
% \be\lb{sminus2}
% \alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} \bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}\bra{\psi_p} \op {\mathbf r}_j \ket{w_p^a}\;.
% \ee
\be\lb{sminus2}
\alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} |\bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}|^2 \;.
\ee
Such value can be straightforwardly compared with results of Eq.~\eqref{staticalpha}, where the localized FS is obtained from GS techniques as
described in Sec.~\ref{FluctuationState}.
This is the well-know $S_{-2}$ sum-rule (see e.g.\cite{Wagner2012}), that expresses the value of the static polarizability
in terms of the molecule's oscillator strengths.
In the context of a numerical evaluation of the excitations, such sum rule can therefore be interpreted as the capability of the pseudo-continuum sector of the excitation to
express the localized FS -- and consequently the reference value of $\alpha_{ij}$ -- at $\omega=0$.

@@

\begin{table}
\begin{tabular}{c|cc||cc}
  Mol. ($N_{at}$) & $\braket{\alpha}_\text{ref}$ & $ \braket{\text{Eq.~\eqref{sminus2}}}$ & \% LS & \% $\Omega_a <$ IP  \\
  \hline
  \hline
  N$_2 (2)$ &12.49 & 12.38 & $1.6\cdot 10^{-6}$ & $1.6\cdot 10^{-6}$ \\
  CO (2)& 15.07 & 14.79 & 8.6 & 8.6 \\
  H$_2$O (3) & 10.87 & 10.72 & 0.1 & 0.1 \\
  Benzene (12)& 71.11 & 70.32 & 3.0 & $7.2 \cdot 10^{-5}$\\
  Aflatoxin (35)& N/A   & 226.27 & 15.87 & 14.84 \\
  C$_{60}$ (60)& 561.59 & 509.60 & 38.47 & 34.41
\end{tabular}\caption{Comparison of the average static polarizabilities for molecules of various sizes (indicated by the number of atoms $N_{at}$) and symmetries, obtained from the reference calculations of Eq.~\eqref{staticalpha} and from the sum over the excitations extracted from the Casida coupling matrix. The last two columns show the percentage of the contribution to Eq.~\eqref{sminus2} which
come from the Localized Sector (LS) of the excitations, by focusing also on the LS before the first IP, where all the excitated states are localized. We see that by increasing the molecules' sizes the LS contribution increases in percentage, however being always largely insufficient; the contribution coming from the pseudocontinuum excitations can never be neglected.}
\end{table}

Such a scenario can be also found in the other molecules we have considered;
varying the molecule size and symmetry does not modify the qualitative behavior of the excitation landscape. In addition the
localized sector of the excitations alone is largely insufficient to express $\alpha_{ij}$.
Such considerations are independent on the number of atoms of the molecule, as the range in $\omega$ needed to
achieve satisfactory agreement is largely above the last valence ionization threshold, which is
an intensive quantity.

Once again, this proves that, for any numerical treatment,  pseudo-continuum excitations are fundamental in view of a correct expression
of the response density below IP.
However, the fulfillement of $S_{-2}$ is not a universal quality indicator of $\chi(\omega)$ as the latter may be still badly represented in the high frequency regime:
it is enough to recall
the strong dependency on the box size of the absorption spectra of Fig.~\ref{CO_exc}.
For all these computational setups, the $S_{-2}$ sum rule is, on the contrary, well satisfied.

@@

From a computational perspective, a set of localized basis functions, which \emph{by design} excludes long-range oscillatory behaviour, may be complete
enough to express $\sop \chi(\omega)$ \emph{below} IP, but it will never be able to capture the entire features of the response density operator above
the ionization threshold: the system's excitations expressed in this basis are reliable \emph{only} in the optical regime.
High-energy excitations belong to a continuum sector (which should not be confused with the continuum eigenstates of the unperturbed GS Hamiltonian)
and the excitation density of states for $\omega >$ IP provided in this way has to be considered as an effective representation of $\sop \chi(\omega)$ in order to fulfill Eq.~\eqref{fsinexc}. This fact has been pointed out several times (see e.g. \cite{giustino2012,giustino2014}) in the literature.

On the contrary, computational treatments that are able to express delocalized states may be such to describe more efficiently the excitations DOS,
especially because of the capability of the basis to capture oscillatory behavior. Like in localized basis sets, excitations below threshold can be
precisely calculated with the same paradigm of traditional GS calculations. However, in this computational setup, to correctly express the linear susceptibility below IP -- more specifically the real part of the fluctuation state -- care should be taken in considering enough delocalized excitations
to guarantee fulfillment of sum-rules like Eq.~\eqref{sminus2}.

%Our results show that \emph{both} the sectors of the excitations energies have to be considered in view of the extraction of computationally reliable quantities, and this even below the first IP.
For generic LR quantities, which are functionals of the linear susceptibility,
the quality of the $\chi(\omega)$ (super)operator will then depend
on the frequency $\omega$ of interest: a given spectral representation may provide high-quality results for static $\omega=0$ regimes, while being
unable to provide convergence for high-energy absorption spectra.
%On the contrary,
Absorption spectra below IP are relatively easy to converge, as they are observables which depend only on localized quantities.
However, the fulfillment of sum rules like the $S_{-2}$ is not \textit{per se}
a guarantee of the quality of high energy absorption spectra.
Moreover, if the basis set employed for the evaluation of the target $\alpha_{ij}$ in  Eq.~\eqref{sminus2} is identical to the one employed for the LR treatment the $S_{-2}$
is trivially satisfied in the basis. It is therefore important to have a set of reference values
in the complete basis set limit.

>>>>

We will see in what follows that the situation is not as simple in the context of LR. Indeed the response density -- even for molecular systems -- looses its localized behavior for high frequencies, independently on the nature of the specific system and perturbation. Furthermore, we will demonstrate that to increase the basis set completeness, delocalized oscillatory degrees of freedom cannot be excluded \textit{a priori}, even in the localized regime.

Lastly, we will show that the reliability of the convergence criteria that can be established for LR -- where no variational theorem exists -- depends explicitly on the target frequency range and on the features of the numerical basis set.

@@

From equation \eqref{rhoPrimeFluctuationStateDef2} we see that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is spatially localized, \textit{i.e.} its evaluation can be restricted on a bounded real-space domain.
Even when such locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strengths $F_j$ small enough to
preserve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behavior of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above these threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;,
\ee
here we have assumed that $\op O$ is a real symmetric operator and chosen a real set of occupied molecular states.
Let us suppose that also the observable of interest $\op O$ can be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.
Once again, it would be enough to consider a local operator for this condition to be met.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime},
realized for $\omega>|\eps_h|$, in which fluctuation states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2}
can be evaluated in a localized domain. In this case a computational description of fluctuation states is only meaningful in ``weak''
sense, where the bound state character of $\bra{\psi_p\op O}$ behaves as a regulator. This is naturally achieved in the above mentioned
Krylov space treatments, which explicitly deal with the expression of the scalar products. However, in this regime, imposing \emph{by design}
a localized behaviour of the FS might reveal very dangerous in view of convergence of the results since the computational basis set has
to be able to express a delocalized FS in the support domain of $\bra{\psi_p\op O}$ .
This fact implies that the degrees of completeness of the computational basis should be increased by adding
the oscillatory degrees of freedom needed to describe the FS in the relevant domain.
Stated otherwise, to achieve precise results in this regime the
underlying basis set has to be able to express delocalized states.

@@

It is possible to reformulate this type of analysis by directly looking at LR quantities that are defined independently from the perturbing operators and target observable. This can be achieved by studying the \emph{susceptibility functional}, defined implicitly through the operator equation
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \left(\omega -\mathcal L \right)^{-1} \commutator{\op \Phi(\omega)}{\dmnot} = \sop{{\cal \chi}}(\omega) \op\Phi(\omega) \;.
\ee
The susceptibility (super)operator $\cal \chi(\omega)$ is therefore closely related to the resolvent of the Liouvillian of Eq.~\eqref{LiouvillianRhopomegaDef1}, which is completely defined
from the molecule Hamiltonian $\hnot[\dm]$.
The ``excitation modes''  of the molecule are defined through the \emph{excitation operators}, satisfying
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$.
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) = \sum_{\{a\}} \,
\frac{\sop B_a}{\omega-\Omega_a} \;,
\ee
where the action of the spectral (super)operator $\sop B_a$ reads
\be\lb{spectralop}
  \sop B_a \cdot = \op E_a \trace{\commutator{\dmnot}{\op{\tilde E}_a} \cdot} \;. \nn
\ee
These formulae allows us to present an excitation-based description of Eq.~\eqref{LinearResponseFunctDef2} as follows
\be\lb{deltaoexc}
  \braket{\delta \op O} =  \sum_{\{a\}} \frac{\trace{\op O \sop B_a \op \Phi(\omega)}}{\omega -\Omega_a}\;.
\ee

@@

To further address this point we exploit the formal relation between response density and susceptibility functional $\dm'(\omega) = \sop \chi(\omega)\op\Phi(\omega)$ and make usage of the representation ~\eqref{rhoPrimeFluctuationStateDef1} of the response density. 
Pursuing this procedure provides the excitation-based representation of FS as follows:
\be\lb{fsinexc}
\ket{f_p^{\mathbf r_j}(\omega)}=
\sum_a \frac{\bra{w_p^a} \op{\mathbf r}_j \ket{\psi_p}}{\omega^2 - \Omega_a^2}
\ket{w_p^a}\;,
\ee
where we have employed the notation $\ket{w_p^a}=\ket{\phi_p^a}+\ket{\chi_p^a}$.
This is the well-known equivalence of the Sternheimer formalism with the spectral representation of the LR susceptibility, that is always valid 
in any computational basis set provided the the full spectrum of excited states (including the pseudo-continuum ones) is considered.
The above equation shows that both the sectors of localized and delocalized states $\ket{w_p^a}$ contribute to the FS for \emph{any} value of $\omega$.
However the discussion of Sec.~\ref{FluctuationState} has shown that the FS is a genuinely localized state below the IP threshold.
The equivalence implies that in this regime the contribution of the delocalized excitations in Eq.~\eqref{fsinexc} has to resum into a localized, asymptotically vanishing FS.
The localization below threshold is manifest for the imaginary part of the FS, that is explicitly written as the sum over $\ket{w_p^a}$ below IP (see Eq.~\eqref{imalpha}). 
Instead, for the real part of FS, the \emph{entire} excitation spectrum contributes to the construction of the FS,
which has to be interpreted as a localized wave-packet in the basis of the pseudo-continuum excitations.

@@

\subsection{Effective representation of $\sop \chi(\omega)$ for localized response densities}

The above arguments have shown that the only quantity which can be expressed in terms of the localized sector of the excitation spectrum is the imaginary part of the susceptibility functional $\sop \chi(\omega)$ in the optical regime (i.e. $\omega <$  IP). The response density for a generic perturbation and $\omega$ value requires the inclusion in the Liouvillian DOS of the delocalized part of the excitation spectrum, even in view of expressing localized fluctuation states.

A given numerical treatment typically provides only an effective description of the sector of delocalized excitation which in turns gives rise to an effective
representation of the susceptibility functional. It appears therefore important to identify indicators that enable us to validate the capability
of $\sop \chi(\omega)$ to express the localized FS (and thus the complete response density $\dm'_\Phi(\omega)$) below the IP threshold.
For instance, the static polarizability tensor $\alpha_{ij}$ can be calculated as
\be\lb{sminus2}
\alpha_{ij} = \sint \dd a \frac{w^a_{ij}}{\Omega_a^2}\,, \quad w^a_{ij} \equiv \sum_{p\in a} |\bra{\psi_p} \op {\mathbf r}_i \ket{w_p^a}|^2 \;.
\ee
Such value can be straightforwardly compared with results of Eq.~\eqref{staticalpha}, where the localized FS is obtained from GS techniques as
described in Sec.~\ref{FluctuationState}.
This is the well-know $S_{-2}$ sum-rule (see e.g.\cite{Wagner2012}), that expresses the value of the static polarizability
in terms of the molecule's oscillator strengths.
In the context of a numerical evaluation of the excitations, such sum rule can therefore be interpreted as the capability of the pseudo-continuum sector of the excitation to
express the localized FS -- and consequently the reference value of $\alpha_{ij}$ -- at $\omega=0$.

@@

\begin{table}
\begin{tabular}{c|cc||cc}
  Mol. ($N_{at}$) & $\braket{\alpha}_\text{ref}$ & $ \braket{\text{Eq.~\eqref{sminus2}}}$ & \% LS & \% $\Omega_a <$ IP  \\
  \hline
  \hline
  N$_2 (2)$ &12.49 & 12.38 & $1.6\cdot 10^{-6}$ & $1.6\cdot 10^{-6}$ \\
  CO (2)& 15.07 & 14.79 & 8.6 & 8.6 \\
  H$_2$O (3) & 10.87 & 10.72 & 0.1 & 0.1 \\
  Benzene (12)& 71.11 & 70.32 & 3.0 & $7.2 \cdot 10^{-5}$\\
  Aflatoxin (35)& N/A   & 226.27 & 15.87 & 14.84 \\
  C$_{60}$ (60)& 561.59 & 509.60 & 38.47 & 34.41
\end{tabular}\caption{Comparison of the average static polarizabilities for molecules of various sizes (indicated by the number of atoms $N_{at}$) and symmetries, obtained from the reference calculations of Eq.~\eqref{staticalpha} and from the sum over the excitations extracted from the Casida coupling matrix. The last two columns show the percentage of the contribution to Eq.~\eqref{sminus2} which
come from the Localized Sector (LS) of the excitations, by focusing also on the LS before the first IP, where all the excited states are localized. We see that by increasing the molecules' sizes the LS contribution increases in percentage, however being always largely insufficient; the contribution coming from the pseudocontinuum excitations can never be neglected.}
\label{alphaTable}
\end{table}

This scenario can be also found for the other molecules presented in Table \ref{alphaTable};
varying the molecule size and symmetry does not modify the qualitative behavior of the excitation landscape. In addition the
localized sector of the excitations alone is largely insufficient to express $\alpha_{ij}$.
Such considerations are independent on the number of atoms of the molecule, as the range in $\omega$ needed to
achieve satisfactory agreement is largely above the last valence ionization threshold, which is
an intensive quantity.

Once again, this proves that, for any numerical treatment,  pseudo-continuum excitations are fundamental in view of a correct expression
of the response density below IP.
However, the fulfillment of $S_{-2}$ is not a universal quality indicator of $\chi(\omega)$ as the latter may be still badly represented in the high frequency regime:
it is enough to recall
the strong dependency on the box size of the absorption spectra of Fig.~\ref{CO_exc}.
For all these computational setups, the $S_{-2}$ sum rule is, on the contrary, well satisfied.

@@

From a computational perspective, a set of localized basis functions, which \emph{by design} excludes long-range oscillatory behavior, may be complete
enough to express $\sop \chi(\omega)$ \emph{below} IP, but it will never be able to capture the entire features of the response density operator above
the ionization threshold: the system's excitations expressed in this basis are reliable \emph{only} in the optical regime.
High-energy excitations belong to a continuum sector (which should not be confused with the continuum eigenstates of the unperturbed GS Hamiltonian)
and the excitation density of states for $\omega >$ IP provided in this way has to be considered as an effective representation of $\sop \chi(\omega)$ in order to fulfill Eq.~\eqref{fsinexc}. This fact has been pointed out several times (see e.g. \cite{giustino2012,giustino2014}) in the literature.

On the contrary, computational treatments that are able to express delocalized states may be adeguate to describe more efficiently the excitations DOS,
especially due to the capability of the basis to capture oscillatory behavior. Like in localized basis sets, excitations below threshold can be
precisely calculated with the same paradigm of traditional GS calculations. However, in this computational setup, to correctly express the linear susceptibility below IP -- more specifically the real part of the fluctuation state -- care should be taken in considering enough delocalized excitations
to guarantee fulfillment of sum-rules like Eq.~\eqref{sminus2}.

For generic LR quantities, which are functionals of the linear susceptibility, the quality of $\sop \chi(\omega)$ will then depend
on the frequency $\omega$ of interest: a given spectral representation may provide high-quality results for static $\omega=0$ regimes, while being
unable to provide convergence for high-energy absorption spectra.
Absorption spectra below IP are relatively easy to converge, as they are observables which depend only on localized quantities.
However, the fulfillment of sum rules like the $S_{-2}$ is not \textit{per se}
a guarantee of the quality of high energy absorption spectra.
Moreover, if the basis set employed for the evaluation of the target $\alpha_{ij}$ in  Eq.~\eqref{sminus2} is identical to the one employed for the LR treatment the $S_{-2}$
is trivially satisfied in the basis. It is therefore important to have a set of reference values
in the complete basis set limit.
